Data Preparation

### 2019-01-09 Data Preparation.Rmd
optional preprocessing (removes stopwords and numbers) from course manuals

### Course Manuals
To expand our topic modeling of course content, we would like to analyse the material in the course manuals. The `corpus_manuals` contains the pdfs of the course manuals for the year 2017-2018 (most recently, available).
```{r}
corpus_manuals <- Corpus(x             = DirSource("Manuals 2017-18"),
                 readerControl = list(reader = readPDF(control = list(text = "-layout"))))
```
## Student Data
The tibble `transcripts` contains the transcripts of a selection of `7` alumni students from the college[^2]. `transcripts` contains information on each course taken by each student such as the receives grade, the semester in which teh course was taken as well as the calendar year.
```{r transcripts}
transcripts <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1kj7tlfJAoN0F7IdoS3saT-2iC3nuDLwnDItNtCKoZ4k/edit#gid=936718979')
print(transcripts)
```

@@

## Course Catalogue
In order to conduct a topic analysis of course content, we extract the *overview* and *description* of each course from the course catalogues.

### Extracting Overviews
We start by extracting the overviews which are 1 or 2 page long and contain important information of a course. To accomplish this, for each catalogue (loop), we first use `grep` to determine which pages should be excluded from the analysis (`pages_to_exclude`). These pages are typically headers. Excluding them before the analysis allows use to keep the code relatively simple. We also make use the fact that the overviews always start with the code of the course, i.e. it starts with one of the elements of `course_code` and use `grep` to identify the first page of each overview. We use grep a 3rd time to identify the page containing the header *Core Coureses (COR)* and which marks the beginning of the overviews in the catalogue, and we loop from this page to the last page of the catalogue.

In the loop, we first determine the content of the current and following page. If the current page is the first page of an overview, then we identify the overview (one or two pages) and save it to the tibble `d_description`. To identify the overview of a course, we check if the following page is either the first page of an overview or a page to exclude. If it is either of these, then the overview correspond to the content of the current page; but if the following page is neiter of these, then the overview correspond to the content of the current page and that of the following page.

```{r overview}
# Setup
n_catalogue <- length(corpus)
calendar_years <- c("2014-2015", "2015-2016", "2016-2017", "2017-2018", "2018-2019")
d_description <- tibble(Code            = character(0),
                        `Calendar Year` = character(0),
                        Overview        = character(0))

course_code <- c("COR", "HUM", "SCI", "SSC", "SKI", "PRO", "UGR", "CAP")
course_code <- paste(course_code, collapse = "|")

content_to_exclude <- c("^Core Courses \\(COR\\)",
                        "^Humanities \\(HUM\\)",
                        "^Sciences \\(SCI\\)",
                        "^Social Sciences \\(SSC\\)",
                        "^Skills Trainings \\(SKI\\)",
                        "^Projects \\(PRO\\)",
                        "^Undergraduate Research \\(UGR\\)",
                        "Appendix")
content_to_exclude <- paste(content_to_exclude, collapse = "|")

# Loop
for(n in 1 : n_catalogue){

  cat <- content(corpus[[n]])

  pages_to_exclude     <- grep(pattern = content_to_exclude, x = cat)
  first_three_letters  <- substr(cat, start = 1, stop = 3)
  first_pages_overview <- grep(pattern = course_code, x = first_three_letters)

  page_start_loop <- grep(pattern = "^Core Courses \\(COR\\)", x = cat) + 1
  page_end_loop   <- length(cat)

  for(page in page_start_loop : page_end_loop){

    current_page           <- page
    following_page         <- page + 1
    current_page_content   <- cat[page          ]
    following_page_content <- cat[following_page]

    if(current_page %in% first_pages_overview){

      if(following_page %in% c(first_pages_overview, pages_to_exclude)) course_overview <- current_page_content
      else course_overview <- paste(current_page_content, following_page_content)

      # Save Overview
      Code <- substring(course_overview, first = 1, last = 7)      
      year <- calendar_years[n]
      d_description <- add_row(d_description,
                               Code            = Code,
                               `Calendar Year` = year,
                               Overview        = course_overview)
    } # close if-statement
  } # close for-loop (page)
} # close for-loop (n)

print(d_description)
```

### Extracting Descriptions
Since the overviews contain a lot of information that does not interest us, we extract the description section of the overviews. The description section contains a brief description (200-500 words) of the content of the course. To accomplish this, for each overview, we need to identify the line where the description starts and the line where it ends. The description section is always preceeded by a header saying *Description* or *Course Description* (`start_descrip`), making it relatively easy to find the starting line of the decription section with `grep`. The description section is usually followed by the literature section which starts with header saying *Literature*, *Recommended Literature* or similar (see `end_descrip`). Using `grep` together with `end_descrip`, we can again identify the ending line of the description section of almost overviews. A handful overviews do not contain a *Literature* section. For these, the section *Instructional Format* marks the end of the description section.
```{r description}
# Setup
symbol_to_line <- c("\r\n", "\n")
symbol_to_line <- paste(symbol_to_line, collapse = "|")

start_descrip <- c("^Description", "^Course Description")
start_descrip <- paste(start_descrip, collapse = "|")

end_descrip   <- c("^Literature$", "^Recommended Literature", "^ Required Litera",
                   "^Required Litera", "^ Literature$", "Literature \\(all")
end_descrip   <- paste(end_descrip, collapse = "|")

d_description <- d_description %>%
  mutate(Description = NA)

# Loop
for(course in 1 : nrow(d_description)){

  overview <- d_description$Overview[course]

  overview_line <- strsplit(x = toString(overview), split = symbol_to_line)[[1]]
  line_start    <- grep(pattern = start_descrip, x = overview_line) + 1
  line_end      <- grep(pattern = end_descrip  , x = overview_line) - 1

  # for overviews w/o Literature section, find section Instructional Format.
  if(is_empty(line_end)){
    line_end <- grep(pattern = "^Instructional format$", x = overview_line) - 1
    print(paste("line_end irregular for", d_description$Code[course],
                "in catalogue", d_description$`Calendar Year`[course]))
    } # close if-statement

  description <- paste(overview_line[line_start : line_end], collapse = " ")

  d_description$Description[course] <- description

print(d_description)
```

### Tidy Text Format
We first create two tibbles `d_overview_tidy` and `d_description_tidy` that respectively store the overviews and the descriptions in the tidy text format, with one row per course-year-word.
```{r tidy text data}
d_overview_tidy <- d_description %>%
  select(Code, `Calendar Year`, Overview) %>%
  unnest_tokens(output = word, input = Overview)
print(d_overview_tidy)

d_description_tidy <- d_description %>%
  select(Code, `Calendar Year`, Description) %>%
  unnest_tokens(output = word, input = Description)
print(d_description_tidy)
```

### Stemming
We then use the function `hunspell__stem`, which returns valid stems for a given word, to stem the words in the tibbles `d_overview_tidy` and `d_description_tidy`. We first create a function `stem_hunspell` which, given a term, returns its most basic stem (the last one from the list of stem returned by `hunspell__stem`). We then apply `stem_hunspell` on the words of `d_overview_tidy`. Since, `stem_hunspell` is not a vectorized function and the number of words in `d_overview_tidy` is large, we first use distinct on the `word` variable to find the a list containing *once* each term present in the course overviews and apply the function `stem_hunspell` on this list. We save the terms and their stems in the tibble `stems`. Since the course descriptions can be conceptualized as a subset of the course overviews, the tibble `stems` contains all the terms present in the descriptions and can therefore safely used to stem the words in `d_description_tidy`.
```{r function stem_hunspell, cache = TRUE}
stem_hunspell <- function(term) {
  # look up the term in the dictionary
  stems <- hunspell_stem(term)[[1]]

  if (length(stems) == 0) { # if there are no stems, use the original term
    stem <- term
  } else { # if there are multiple stems, use the last one
    stem <- stems[[length(stems)]]
  }

  return(stem)
}

stems <- d_overview_tidy %>% 
  select(word) %>%
  distinct %>%
  arrange(word) %>%
  mutate(word_stem = purrr::map_chr(.x = word,
                                    .f = stem_hunspell))

# terms for which the stem differs from the original word
stems %>%
  filter(word != word_stem)
```

```{r stemming}
d_overview_stem <- d_overview_tidy %>%
  left_join(stems, by = "word") %>%
  rename(word_original = word,
         word = word_stem) %>%
  filter(!word %in% stop_words$word)
print(d_overview_stem) # See humanities - humanity

d_description_stem <- d_description_tidy %>%
  left_join(stems, by = "word") %>%
  rename(word_original = word,
         word = word_stem) %>%
  filter(!word %in% stop_words$word)
print(d_description_stem) # See starting - start
```

## Course Manual 
We create a dataframe from the Corpus:
```{r}
manuals_td <- tidy(corpus_manuals)
```

From `manuals_td` we are only interested in the text, and the Code of the course. However, we don't have the code but we have a document id that includes it. Therefore we select only the `text` and `id` columns of the `manuals_td`. Then, we extract the code from the `id` and save it in under `Code`, We aso extract the concentration from the `id` and save it under `Concentration`. Finally, we unnest the dataframe to have only one word per row, and call this ``manuals_tokens.

```{r}
manuals_tokens <- manuals_td %>%
  select(id,text) %>%
  mutate(Concentration= substring(id, 1, 3), Code= substring(id,1,7))%>%
  unnest_tokens(word, text)
```

# Save Data

>>>>

### Course Manuals
To expand our topic modeling of course content, we would like to analyse the material in the course manuals. The `corpus_manuals` contains the pdfs of the course manuals for the year 2017-2018 (most recently, available).
```{r}
corpus_manuals <- Corpus(x             = DirSource("Manuals 2017-18"),
                 readerControl = list(reader = readPDF(control = list(text = "-layout"))))
```
## Student Data
The tibble `transcripts` contains the transcripts of a selection of `7` alumni students from the college[^2]. `transcripts` contains information on each course taken by each student such as the receives grade, the semester in which teh course was taken as well as the calendar year.
```{r transcripts}
transcripts <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1kj7tlfJAoN0F7IdoS3saT-2iC3nuDLwnDItNtCKoZ4k/edit#gid=936718979')
print(transcripts)
```
COMMENT SOFIA: save a_clean_transcript as d_transcript

@@

## Course Manual 
We create a dataframe from the Corpus:
```{r}

manuals_td <- tidy(corpus_manuals)

```

From `manuals_td` we are only interested in the text, and the Code of the course. However, we don't have the code but we have a document id that includes it. Therefore we select only the `text` and `id` columns of the `manuals_td`. Then, we extract the code from the `id` and save it in under `Code`, We aso extract the concentration from the `id` and save it under `Concentration`. Finally, we unnest the dataframe to have only one word per row, and call this `manuals_tokens`, this creates a variable `word` and adds all of the text from the course manuals under this column.

```{r}
manuals_tokens <- manuals_td %>%
  select(id,text) %>%
  mutate(Concentration= substring(id, 1, 3), Code= substring(id,1,7))%>%
  unnest_tokens(word, text)
```
###(Optional) Pre-processing Course Manuals- Removing stopwords
Some characters appear with sufficient frequency to be relevant but not often enough accross all courses to be deleted when using tf-idf (see sections bellow), it is therefore convenient to add some stopwords and remove numbers. For this we import stop_words from tidytext, and search all our words with grep for stand alone numbers. We save these words in different dataframes and then exclude with an antijoin.
```{r}

data("stop_words")  #imports common stopwords

my_stop_words <- data.frame (word = c(1:11, "students", "______________________________________________________________________________")) #no stemming ("students")
number_stopwords <- number_stopwords <- manuals_tokens %>% #can be cleaner by using unique (?)
  mutate(is_number=grepl("^[-]{0,1}[0-9]{0,}.{0,1}[0-9]{1,}$",word))%>% #finds stand alone numbers (incl. -, and .)
  filter(is_number==T)

manuals_tokens_preprocessed <- manuals_tokens %>%
  anti_join(stop_words)     %>%
  anti_join(my_stop_words) %>%
  anti_join(number_stopwords, by = c("word"="word"))

```
# Save Data
